---
title: "Tests: Edgar Anderson's Iris Data"
author: "Brandon M. Greenwell"
output: html_document
---

```{r setup}
# Load required packages
library(ggplot2)
library(grid)
library(pdp)

# Test function
iris_test <- function(object, ...) {
  trn <- subset(iris, select = -Species)
  pd1 <- partial(object, pred.var = "Petal.Length", which.class = 1, 
                 prob = TRUE, train = trn, ...)
  pd2 <- partial(object, pred.var = "Petal.Length", which.class = 2, 
                 prob = TRUE, train = trn, ...)
  pd3 <- partial(object, pred.var = "Petal.Length", which.class = 3, 
                 prob = TRUE, train = trn, ...)
  pd <- rbind(cbind(pd1, "Class" = "1"), cbind(pd2, "Class" = "2"),
              cbind(pd3, "Class" = "3"))
  ggplot(pd, aes(x = Petal.Length, y = yhat, color = Class)) +
    geom_line() +
    xlab("Petal length") +
    ylab("Probability") +
    ggtitle(deparse(substitute(object))) +
    theme_light()
}
```


# Discriminant analysis


## Package: `MASS`

```{r mass_lda}
iris_lda <- MASS::lda(Species ~ ., data = iris)
iris_test(iris_lda)
```

```{r mass_qda}
iris_qda <- MASS::qda(Species ~ ., data = iris)
iris_test(iris_qda)
```


## Package: `mda`

```{r mda_fda}
iris_fda <- mda::fda(Species ~ ., data = iris)
iris_test(iris_fda)
```

```{r mda_mda}
iris_fda <- mda::mda(Species ~ ., data = iris)
iris_test(iris_fda)
```


# Decision trees


## Package: `rpart`

```{r rpart_rpart}
iris_rpart <- rpart::rpart(Species ~ ., data = iris)
iris_test(iris_rpart)
```


## Package: `C50`

```{r C50_C50}
set.seed(101)  # for reproducibility
iris_C5.0 <- C50::C5.0(Species ~ ., data = iris, trials = 100)
iris_test(iris_C5.0)
```


## Package: `party`

```{r party_ctree}
iris_ctree <- party::ctree(Species ~ ., data = iris)
iris_test(iris_ctree)
```


## Package: `partykit`

```{r partykit_ctree}
iris_ctree2 <- partykit::ctree(Species ~ ., data = iris)
iris_test(iris_ctree2)
```


# Bagging


## Package: `adabag`

```{r adabag_bagging}
set.seed(101)  # for reproducibility
iris_bagging <- adabag::bagging(Species ~ ., data = iris)
iris_test(iris_bagging)
```


## Package: `ipred`

```{r ipred_bagging}
set.seed(101)  # for reproducibility
iris_ipred <- ipred::bagging(Species ~ ., data = iris, nbagg = 500)
iris_test(iris_ipred)
```


# Random forests


## Package: `randomForest`

```{r randomForest_randomForest}
set.seed(101)  # for reproducibility
iris_rf <- randomForest::randomForest(Species ~ ., data = iris)
iris_test(iris_rf)
```


## Package: `party`

```{r party_cforest}
set.seed(101)  # for reproducibility
iris_crf <- party::cforest(Species ~ ., data = iris)
iris_test(iris_crf, quantiles = TRUE)
```


## Package: `partykit`

```{r partykit_cforest}
set.seed(101)  # for reproducibility
iris_crf2 <- partykit::cforest(Species ~ ., data = iris)
iris_test(iris_crf2, quantiles = TRUE)
```


## Package: `ranger`

```{r ranger_ranger}
set.seed(101)  # for reproducibility
iris_ranger <- ranger::ranger(Species ~ ., data = iris, probability = TRUE)
iris_test(iris_ranger)
```


# Boosting


## Package: `adabag`

```{r adabag_boosting}
set.seed(101)  # for reproducibility
iris_boosting <- adabag::boosting(Species ~ ., data = iris)
iris_test(iris_boosting)
```


## Package: `gbm`

```{r gbm_gbm}
set.seed(101)  # for reproducibility
iris_gbm <- gbm::gbm(Species ~ ., 
                     data = iris,
                     distribution = "multinomial",
                     n.trees = 5000,
                     interaction.depth = 3,
                     shrinkage = 0.001,
                     cv.folds = 5,
                     verbose = FALSE)
best.iter <- gbm::gbm.perf(iris_gbm, method = "cv", plot.it = FALSE)
iris_test(iris_gbm, n.trees = best.iter)
```


## Package: `xgboost`

```{r xgboost_xgboost}
set.seed(101)  # for reproducibility
iris_xgb <- xgboost::xgboost(
  data = data.matrix(subset(iris, select = -Species)), 
  label = unclass(iris$Species) - 1, num_class = 3, 
  objective = "multi:softprob", nrounds = 100, max_depth = 3, eta = 0.1, 
  gamma = 0, colsample_bytree = 0.8, min_child_weight = 1, subsample = 0.7,
  verbose = 0
)
iris_test(iris_xgb)
```


# Neural networks

## Package: `nnet`

```{r nnet_nnet}
set.seed(101)  # for reproducibility
iris_nnet <- nnet::nnet(Species ~ ., data = iris, size = 10, decay = 0.1, 
                        maxit = 500, trace = FALSE)
iris_test(iris_nnet)
```


# Support vector machines


## Package: `e1071`

```{r e1071_svm}
iris_svm <- e1071::svm(Species ~ ., data = iris, type = "C-classification", 
                       probability = TRUE)
iris_test(iris_svm)
```


## Package: `kernlab`

```{r kernlab_ksvm}
iris_ksvm <- kernlab::ksvm(Species ~ ., data = iris, type = "C-svc", 
                           prob.model = TRUE)
iris_test(iris_ksvm)
```


# Linear/generalized linear models


## Package: `nnet`

```{r nnet_multinom}
set.seed(101)  # for reproducibility
iris_multinom <- nnet::multinom(Species ~ ., data = iris, trace = FALSE)
iris_test(iris_multinom)
```
