---
title: "Making pdp faster"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Making pdp faster}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

Computing partial dependence is embarrassingly simple but potentially
expensive: for each point in the grid of predictor values, the model has to
score a modified copy of the entire training set. This vignette covers the
options **pdp** provides for speeding that up.

```{r boston-rf}
library(pdp)
library(randomForest)

data(boston)
set.seed(101)
boston.rf <- randomForest(cmedv ~ ., data = boston, ntree = 250)
```

## Batched predictions (`batch.size`)

By default, `partial()` calls `predict()` once per grid point. Most
`predict()` methods have non-trivial per-call overhead, so it is often much
faster to *stack* several grid points together and score them in a single
call. The `batch.size` argument specifies the (approximate) maximum number of
rows to score per call:

```{r batch}
system.time(  # classic: one predict() call per grid point
  pd1 <- partial(boston.rf, pred.var = "lstat", train = boston)
)
system.time(  # batched: score up to one million rows per predict() call
  pd2 <- partial(boston.rf, pred.var = "lstat", train = boston,
                 batch.size = 1e6)
)
identical(pd1, pd2)
```

The trade-off is memory: with `batch.size = 1e6`, up to a million rows are
held in memory at once. Pick a batch size that comfortably fits your machine.
Note that `batch.size` requires the prediction function to return one
prediction per row of `newdata`, so it cannot be combined with a `pred.fun`
that aggregates its own predictions.

## Parallel processing

`partial()` can compute the grid points (or batches) in parallel via the
[foreach](https://cran.r-project.org/package=foreach) package. Register a
parallel backend (e.g., with **doParallel**) and set `parallel = TRUE`:

```{r parallel, eval=FALSE}
library(doParallel)

cl <- makeCluster(4)  # use 4 workers
registerDoParallel(cl)
pd <- partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE,
              train = boston, parallel = TRUE)
stopCluster(cl)
```

If the model's `predict()` method lives in a package, pass it along via
`paropts` (e.g., `paropts = list(.packages = "ranger")`) so the workers can
find it.

## The recursive method for gbm

For **gbm** models, `partial()` defaults to `recursive = TRUE`, which uses
Friedman's weighted tree traversal method (implemented in C++) instead of the
brute force approach. It is *much* faster since it never has to touch the
training data:

```{r gbm}
library(gbm)

set.seed(103)
boston.gbm <- gbm(cmedv ~ ., data = boston, distribution = "gaussian",
                  n.trees = 500, interaction.depth = 3, shrinkage = 0.1)

system.time(
  pd.recursive <- partial(boston.gbm, pred.var = "lstat", n.trees = 500,
                          train = boston)  # recursive = TRUE is the default
)
system.time(
  pd.brute <- partial(boston.gbm, pred.var = "lstat", n.trees = 500,
                      recursive = FALSE, train = boston, batch.size = 1e6)
)
```

The two methods agree closely (they handle weights slightly differently), but
note that the recursive method cannot be used with `ice = TRUE`, `pred.fun`,
or `inv.link`.

## Fast approximate PDPs (`approx = TRUE`)

Finally, `approx = TRUE` computes a much cheaper *approximation* to partial
dependence: rather than averaging over the training data, predictions are
computed for a single "exemplar" observation (continuous predictors are fixed
at their median; categorical predictors at their most frequent value). This is
in the same spirit as the **plotmo** package and is most useful as a quick
first look at very large data sets:

```{r approx}
partial(boston.rf, pred.var = "lstat", approx = TRUE, plot = TRUE,
        plot.engine = "tinyplot", train = boston)
```

The underlying `exemplar()` function is also exported; you can achieve the
same thing (with more control) by passing an exemplar record to `train` and,
optionally, your own `pred.grid`.
