---
title: "Single-variable PDPs with standard deviations"
author: "Brandon M. Greenwell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: pdp.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "left"
)
```

It is possible to construct PDPs with pointwise variability estimates (e.g., +/- one standard deviation). This can be done easily with the `pred.fun` argument of `partial()`. To illustrate, we'll fit a random forest to the famous iris data and construct a PDP for the most important variable that displays pointwise standard deviation bands. 

In the code chunk below, we use the **ranger** package [@R-ranger] to fit a random forest to the built-in iris data. Note that in order to use the **vip** [@R-vip] and **pdp** [@R-pdp] packages for classification models, we need to be able to compute the predicted class probabilities. Therefore, we must specify `probability = TRUE` in the call to `ranger()`.

```{r iris-01}
library(ranger)  # for fitting random forests
set.seed(101)  # for reproducibility
rfo <- ranger(Species ~ ., data = iris, probability = TRUE, 
              importance = "impurity")
```

Next, we'll use the **vip** package to construct a variable importance plot (VIP) from the fitted random forest. The default for `vip()` is to construct model-specific VIPs if availble. Since we specified `importance = "impurity"` in the call to `ranger()`, `vip()` will plot the impurity-based variable importance scores common with most decision tree-based algorithms.

```{r iris-02, fig.width=6, fig.asp=0.618, out.width="70%", fig.cap="**Figure 1** Variable importance plot from a random forest fit to the iris data set."}
library(vip)  # for variable importance plots
vip(rfo)
```

It appears from **Figure 1** that `Petal.Width` and `Petal.Length` are the most important features in predicting `Species`. Now that we've identified these important features, we'll construct PDPs for each that include pointwise standard deviation bands. To do this, we'll specify a special prediction function that returns three components: the average prediction, the average prediction minus one standard deviation, and the average prediction plus one standard deviation. This function is defined below; note that this function requires the arguments `object` and `newdata`.

```{r iris-03}
pred_wrapper <- function(object, newdata) {
  p <- predict(object, data = newdata)$predictions[, 1L, drop = TRUE]
  c("avg" = mean(p), "avg-1sd" = mean(p) - sd(p), "avg+1sd" = mean(p) + sd(p))
}
```

Next, we just supply this function via the `pred.fun` argument in the call to `partial()`:

```{r iris-04}
library(pdp)  # for partial dependence plots
pd1 <- partial(rfo, pred.var = "Petal.Width", pred.fun = pred_wrapper)
pd2 <- partial(rfo, pred.var = "Petal.Length", pred.fun = pred_wrapper)
```

To plot the resulting PDPs, we could've just specified `plot = TRUE` in the previous calls to `partial()`. However, since PDPs can be computationally expensive to compute (though, not in this example), it is good practice to store the results first, and then manually construct the plot. We'll use the convenient `autoplot()` function provided by **pdp** to construct the plots (which requires that the **ggplot2** package [@R-ggplot2] be laoded first). The results are displayed in **Figure 2**.

```{r iris-05, fig.width=7, fig.height=3.5, out.width="100%", fig.cap="**Figure 2** Partial dependence of `Species` on `Petal.Width` (left) and `Petal.Length` (right). The red lines is the PDP while the black lines indicate pointwise +/- one standard deviation."}
library(ggplot2)  # for autoplot() generic
pdp1 <- autoplot(pd1) + 
  theme_light() +
  labs(x = "Petal width (mm)", y = "Partial dependence") +
  theme(legend.position = "none")
pdp2 <- autoplot(pd2) + 
  theme_light() +
  labs(x = "Petal length (mm)", y = "Partial dependence") +
  theme(legend.position = "none")
grid.arrange(pdp1, pdp2, nrow = 1)  # display plots side by side
```
